---
title: "Survival Analysis - Miniproject"
author: "fib 4107"
date: "Autumn 2020"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(survival)
library(readr)
andersen.plot=function(surv,log_scale=FALSE){
    #surv is survfit object based on stratified Cox ph model

    nstrata=surv$strata
    if (log_scale) {
        surv$cumhaz = surv$cumhaz[surv$cumhaz != 0] 
        plot(c(0,0),c(0,0),xlim=c(min(log(surv$cumhaz)),max(log(surv$cumhaz))),ylim=c(min(log(surv$cumhaz)),max(log(surv$cumhaz))),xlab="H1",ylab="H2",type="n",main='Log-scale')    
    } else {
        #plot(c(0,0),c(0,0),xlim=c(min(surv$cumhaz),max(surv$cumhaz)),ylim=c(min(surv$cumhaz),max(surv$cumhaz)),xlab="H1",ylab="H2",type="n")
        plot(c(0,0),c(0,0),xlim=c(0,max(surv$cumhaz)),ylim=c(0,max(surv$cumhaz)),xlab="H1",ylab="H2",type="n")
    }
    
    count=1
    for (i in 1:length(nstrata))
        for (j in i:length(nstrata))
    # for (i in 1:1)
    #     for (j in 2:2)
            if (i != j){


                if (i==1) times1=surv$time[1:nstrata[1]] else times1=surv$time[sum(nstrata[1:(i-1)]+1):sum(nstrata[1:i])]

                if (j==1) times2=surv$time[1:nstrata[1]] else times2=surv$time[sum(nstrata[1:(j-1)]+1):sum(nstrata[1:j])]


                #print(times1)
                #print(times2)
                mintime=min(c(max(times1),max(times2)))
                times=sort(unique(c(times1,times2)))
                times=times[times<=mintime]
                curves=summary(surv,times=times)
                ntimes=length(times)
                #print(times)
                
                if (log_scale) {
                    temp1=curves$cumhaz[c(1:ntimes)+(i-1)*ntimes]
                    temp2=curves$cumhaz[c(1:ntimes)+(j-1)*ntimes]
                    temp1_z = temp1 != 0
                    temp2_z = temp2 != 0
                    temp1=temp1[temp1_z & temp2_z]
                    temp2=temp2[temp1_z & temp2_z]
                } else {
                    temp1=curves$cumhaz[c(1:ntimes)+(i-1)*ntimes]
                    temp2=curves$cumhaz[c(1:ntimes)+(j-1)*ntimes]
                }
                
                #print(ntimes)
                #print(summary(curves))
                #print(curves$time)
                #print(curves$cumhaz)
                #print(temp1)
                #print(temp2)
                #include=temp1 > 0 & temp2 >0
                #temp1=temp1[include]
                #temp2=temp2[include]
                if (log_scale) {
                    points(log(temp1),log(temp2),xlab="",ylab="",col=count)
                    fit=lm(log(temp2)~log(temp1))
                } else {
                    points(temp1,temp2,xlab="",ylab="",col=count)
                    fit=lm(temp2~-1+temp1)
                }
                abline(fit,col=count)
                count=count+1
            }
}
```

# Importing the Data
This miniproject will concern a cirrhosis dataset.
The dataset is imported with the following code.
We have hosted a copy of the dataset in a GitHub repository \texttt{Rosenkrands/survival-analysis-miniproject}.

```{r}
#data_path = 'https://raw.githubusercontent.com/Rosenkrands/survival-analysis-miniproject
#/main/cirrhosis.dat'
#data_path = 'https://people.math.aau.dk/~rw/Undervisning/DurationAnalysis/Data/cirrhosis.dat'
#data = read.table(data_path)
data <- read_table2("cirrhosis.csv", 
    col_names = FALSE)
names(data) <- c('pat_id','cens','obs_t','treat','sex','ascites','age','prothrombin')
```

# Effect of Prednison
The first task of the miniproject is to investigate the effect of prednison by considering estimates of the survival functions for the treatment groups and by performing a log-rank test.

## Comparison of Survival Functions
First we will plot the estimated survival functions for the two treatment groups.

```{r}
treat_groups <- split(data, data$treat)
names(treat_groups) <- c('prednisone','placebo')
predni_fit <- survfit(Surv(treat_groups[['prednisone']]$obs_t,
                           treat_groups[['prednisone']]$cens)~1,type='kaplan-meier') 
placebo_fit <- survfit(Surv(treat_groups[['placebo']]$obs_t,
                            treat_groups[['placebo']]$cens)~1,type='kaplan-meier')
```

```{r, echo=F, fig.height=3, fig.width=6}
par(mfrow=c(1,2))
plot(predni_fit,main='Prednisone')
plot(placebo_fit,main='Placebo')
```

## Log-Rank Test
Next we can consider differences between the two groups and we conclude the effect of prednison is very insignificant from the below summary.

```{r, echo=F}
survdiff(Surv(data$obs_t,data$cens) ~ data$treat)
```

# Cox Proportional Model

The second task is to fit a cox proportional model to the dataset.

```{r}
coxph_model = coxph(Surv(data$obs_t,data$cens) ~ data$treat + data$sex + data$ascites + data$age + data$prothrombin)
```

## Model Assesment

```{r}
fit.treat = coxph(Surv(data$obs_t,data$cens) ~ data$treat + data$sex + data$ascites + data$age + data$prothrombin)
devres=residuals(fit.treat,type='deviance')
par(mfrow=c(1,2))
hist(devres)
boxplot(devres~data$cens)
```

For each of the variables we fit a stratified model and asses the effects.

### Treatment
First we look at the treatment variable, i.e. whether or not the patient were treated with the placebo or prednison.

```{r}
fit.treat = coxph(Surv(data$obs_t,data$cens) ~ strata(data$treat) + data$sex + data$ascites + data$age + data$prothrombin)
surv.treat = survfit(fit.treat)
par(mfrow=c(1,2))
andersen.plot(surv.treat)
andersen.plot(surv.treat,log_scale = T)
```
We conclude that the points fit reasonably well with the assumption of proportional hazard, with slope 1 which indicates that there are no difference in hazard between the two groups.

### Sex
Next we look at the sex variable.

```{r}
fit.treat = coxph(Surv(data$obs_t,data$cens) ~ data$treat + strata(data$sex) + data$ascites + data$age + data$prothrombin)
surv.treat = survfit(fit.treat)
par(mfrow=c(1,2))
andersen.plot(surv.treat)
andersen.plot(surv.treat,log_scale=T)
```

The plots indicate the that hazards are proportional, but unlike with the treatment variable, there seems to be a different hazard rate between the two groups as the slope is approximately 2. 

### Ascites
The ascites variable are again of categorical nature so we can construct andersen plots.

```{r}
fit.treat = coxph(Surv(data$obs_t,data$cens) ~ data$treat + data$sex + strata(data$ascites) + data$age + data$prothrombin)
surv.treat = survfit(fit.treat)
par(mfrow=c(1,2))
andersen.plot(surv.treat)
andersen.plot(surv.treat,log_scale=T)
```

On the regular plot the hazard rates seem to be proportional, however when looking on the plot with log scales there are some outliers for the black category.

### Age

```{r}
fit.treat = coxph(Surv(data$obs_t,data$cens) ~ data$treat + data$sex + data$ascites + data$age + data$prothrombin)
devres=residuals(fit.treat,type='deviance')
par(mfrow=c(1,1))
plot(data$age,devres)
```

Looks good, fewer point for lower ages as not many observations.

### Prothrombin

```{r}
par(mfrow=c(1,2))
plot(data$prothrombin,devres)
hist(data$prothrombin)
```

Looks good, many at 100 as this is the normal level. Many below; old people mostly maybe related to illness.

# Hypothesis test
Summary from cox proportional model.

```{r, echo=F}
coxph_model
```

We observe that the only insignificant variable is the type of treatment, prednison or placebo.

Therefore we accept the null hypothesis of no effect of prednison treatment.

# Transformations

## Age
martingale residuals for age

```{r}
fitnoage = coxph(Surv(data$obs_t,data$cens) ~ data$treat + data$sex + data$ascites + data$prothrombin)
resall=residuals(fitnoage)
par(mfrow=c(1,1))
scatter.smooth(data$age,resall)
```

relation looks linear, indicating that including as continuous variable is good.

## Prothrombin
martingale residuals for prothrombin.

```{r}
fitnopro = coxph(Surv(data$obs_t,data$cens) ~ data$treat + data$sex + data$ascites + data$age)
resall=residuals(fitnopro)
par(mfrow=c(1,1))
scatter.smooth(data$prothrombin,resall)
```

relation could suggest discretization at 55 percent would benefit the model.

```{r}
fit.pro <- coxph(Surv(data$obs_t,data$cens) ~ data$treat + data$sex + data$ascites + data$age + cut(data$prothrombin, breaks=c(0,55,150)))
devres_pro=residuals(fit.pro,type='deviance')
par(mfrow=c(2,2))
hist(devres)
boxplot(devres~data$cens)
hist(devres_pro)
boxplot(devres_pro~data$cens)
```

however looking at results it is not worth the added complexity.

# Interaction

Model with interaction terms.

```{r}
fit = coxph(Surv(data$obs_t,data$cens) ~ data$treat + data$sex + data$ascites + data$age + data$prothrombin + data$treat*(data$sex + data$ascites + data$age + data$prothrombin))
fit
```

model indicates significant interaction between treatment and ascites variable.

# Estimate survival function
Estimate and plot survival function.

```{r}
fit = coxph(Surv(obs_t,cens) ~ treat + sex + ascites + age + prothrombin + treat*ascites, data=data)
newdata = data.frame(treat=c(0),sex=c(1),ascites=c(1),age=c(57),prothrombin=c(85))
surv_fit = survfit(formula=fit,newdata=newdata)
plot(surv_fit)
```

Day number 2928 is the first observation time after 8 years.

```{r}
surv_fit$surv[surv_fit$time == 2928]
```

poor guy.
